---
title: "Raster Summarization"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "90%",
  dpi = 150,
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

Comparing R `raster` and `gdal-summarize.py`, which uses Python with GDAL bindings, for aggregating raster data across layers.

## Testing accuracy

To make sure `gdal-summarize.py` is doing what we think it's doing, let's apply it to the GeoTIFF `data/small1.tif`. This tiny file stores two raster bands with dimensions 3x3 and the cells contains positive integers, zeros, and missing values. We'll compare the results of `gdal-summarize.py` and the equivalent using the `raster` R package.

```{r test-data}
library(raster)
library(rasterVis)
library(viridis)
library(ebirdst)
library(tidyverse)

r <- stack("data/small1.tif")
as.matrix(r[[1]])
levelplot(r, 
          col.regions = viridis::viridis,
          xlab = NULL, ylab = NULL, 
          scales = list(draw = FALSE),
          names.attr = c("Band 1", "Band 2"))
```

Let's start by summarizing these data in Python using `gdal-summarize.py` with the four different summary functions: sum, mean, count, and richness. We'll summarize across all (2) bands.

```{r test-py}
# prepare the commands
funs <- c("sum", "mean", "count", "richness")
cmd <- str_glue("./gdal-summarize.py data/small1.tif -w -q ",
                "-f {funs} -o output/small1_{funs}.tif")
# run the commands
walk(cmd, ~ {print(.); system(paste("source ~/.bash_profile;", .))})
# read in the files
s_python <- str_glue("output/small1_{funs}.tif") %>% 
  stack() %>% 
  setNames(funs)
```

Now let's do the same thing in R using `raster`.

```{r test-raster}
r_sum <- calc(r, sum, na.rm = TRUE)
r_mean <- calc(r, mean, na.rm = TRUE)
r_count <- calc(r >= 0, sum, na.rm = TRUE)
r_richness <- calc(r > 0, sum, na.rm = TRUE)
# stack the raster results
s_raster <- stack(r_sum, r_mean, r_count, r_richness) %>% 
  setNames(c("sum", "mean", "count", "richness"))
```

Now we can compare the results. For example, the results for the mean of the layers look identical. Noticing that the zero, non-zero, and missing values are all matching.

```{r test-compareone}
print("Python sum: ")
as.matrix(s_python[["mean"]])
print("R sum:")
as.matrix(s_raster[["mean"]])
```

Let's compare across all four summary functions using `compareRaster()`, which checks whether two `Raster` objects are identical in extent, dimensions, projection, resolution, and values.

```{r test-compare}
for (f in funs) {
  comp <- compareRaster(s_python[[f]], s_raster[[f]], 
                        res = TRUE, values = TRUE,
                        stopiffalse = FALSE)
  print(str_glue("{f} rasters identical: {ifelse(comp, 'Yes', 'No')}"))
}
```

### Alternate usage

`gdal-summarize.py` can be used in a variety of other ways. Here we test to make sure a couple of the other ways of using it work as expected.

- Summarize the first band across a set of raster files

```{r test-alt1}
# raster
r1 <- stack("data/small1.tif")
r2 <- stack("data/small2.tif")
r12 <- stack(r1[[1]], r2[[1]]) %>% 
  calc(fun = mean, na.rm = TRUE)

# python
f_tmp <- "output/small_temp.tif"
str_glue("source ~/.bash_profile;",
         "./gdal-summarize.py data/small1.tif data/small2.tif -w -q ",
         "-f mean -o {f_tmp}") %>% 
  {print(.); system(.)}
p12 <- raster(f_tmp)
tibble(raster = r12[], python = p12[])
```

- Summarize a defined band across a set of raster files

```{r test-alt2}
# raster
r12 <- stack(r1[[2]], r2[[2]]) %>% 
  calc(fun = mean, na.rm = TRUE)

# python
str_glue("source ~/.bash_profile;",
         "./gdal-summarize.py data/small1.tif data/small2.tif -w -q ",
         "-b 2 -f mean -o {f_tmp}") %>% 
  {print(.); system(.)}
p12 <- raster(f_tmp)
tibble(raster = r12[], python = p12[])
```

- Summarize different bands across each raster file, band 1 for the first file and band 2 for the second

```{r test-alt3}
# raster
r12 <- stack(r1[[1]], r2[[2]]) %>% 
  calc(fun = mean, na.rm = TRUE)

# python
str_glue("source ~/.bash_profile;",
         "./gdal-summarize.py data/small1.tif data/small2.tif -w -q ",
         "-b 1 2 -f mean -o {f_tmp}") %>% 
  {print(.); system(.)}
p12 <- raster(f_tmp)
tibble(raster = r12[], python = p12[])
```

```{r cleanup, echo=FALSE}
unlink(f_tmp)
```

## Efficiency

Now that we've confirmed everything is working, let's use one of the larger example datasets to compare the efficiency of these approaches. This file stores 9 raster bands with dimensions 250x250 and the cells contains positive integers, zeros, and missing values. We'll compare the processing speed of `gdal-summarize.py` and the equivalent using the `raster` R package.

```{r eff-python}
# prepare the commands
funs <- c("sum", "mean", "count", "richness")
cmd <- str_glue("source ~/.bash_profile;",
                "./gdal-summarize.py data/large1.tif -w -q ",
                "-f {funs} -o output/large1_{funs}.tif")
# run the commands
p_results <- tibble(fun = funs, cmd = cmd) %>% 
  mutate(time_python = map_dbl(cmd, ~ system.time(system(.))[3])) %>% 
  select(-cmd)
p_stack <- str_glue("output/large1_{p_results$fun}.tif") %>% 
  stack() %>% 
  setNames(p_results$fun) %>% 
  setMinMax()
plot(p_stack, col = viridis(100))
```

Now let's do the same thing in R using `raster`.

```{r eff-raster}
r <- stack("data/large1.tif") %>% readAll()
t_sum <- system.time({r_sum <- calc(r, sum, na.rm = TRUE)})[3]
t_mean <- system.time({r_mean <- calc(r, mean, na.rm = TRUE)})[3]
t_count <- system.time({r_count <- calc(r >= 0, sum, na.rm = TRUE)})[3]
t_richness <- system.time({r_richness <- calc(r > 0, sum, na.rm = TRUE)})[3]
# stack the raster results
r_results <- tibble(fun = c("sum", "mean", "count", "richness"), 
                    time_raster = c(t_sum, t_mean, t_count, t_richness))
r_stack <- stack(r_sum, r_mean, r_count, r_richness) %>% 
  setNames(c("sum", "mean", "count", "richness"))
plot(r_stack, col = viridis(100))
```

The results look quite similar, but let's take the difference between the Python and R versions to check.

```{r eff-compare}
difference <- abs(r_stack - p_stack)
cellStats(difference, max)
cellStats(difference, mean)
plot(difference)
```

The count and richness summaries are identical, while the sum and mean summaries have small differences in the 7th decimal place, likely the result of floating point math. Finally, let's look at the difference in run times between the methods.

```{r eff-time}
inner_join(p_results, r_results, by = "fun")
```

# eBird Status & Trends

Finally, applying this to some real data from eBird Status & Trends. First, summarizing with Python.

```{r ebirdst-py, eval = FALSE}
f_abd <- "data/woothr.tif"
abd <- stack(f_abd)

# python
funs <- c("sum", "mean", "count", "richness")
cmd <- str_glue("source ~/.bash_profile;",
                "./gdal-summarize.py {f_abd} -w -q ",
                "-f {funs} -o output/woothr_{funs}.tif")
# run the commands
p_results <- tibble(fun = funs, cmd = cmd) %>% 
  mutate(time_python = map_dbl(cmd, ~ system.time(system(.))[3])) %>% 
  select(-cmd)
p_stack <- str_glue("output/woothr_{p_results$fun}.tif") %>% 
  stack() %>% 
  setNames(p_results$fun) %>% 
  setMinMax()
```

And now with the `raster` R package.

```{r ebirdst-r, eval = FALSE}
t_sum <- system.time({r_sum <- calc(abd, sum, na.rm = TRUE)})[3]
t_mean <- system.time({r_mean <- calc(abd, mean, na.rm = TRUE)})[3]
t_count <- system.time({r_count <- calc(abd >= 0, sum, na.rm = TRUE)})[3]
t_richness <- system.time({r_richness <- calc(abd > 0, sum, na.rm = TRUE)})[3]
# stack the raster results
r_results <- tibble(fun = c("sum", "mean", "count", "richness"), 
                    time_raster = c(t_sum, t_mean, t_count, t_richness))
r_stack <- stack(r_sum, r_mean, r_count, r_richness) %>% 
  setNames(c("sum", "mean", "count", "richness"))
```

Comparing the results.

```{r ebirdst-compare, eval = FALSE}
inner_join(p_results, r_results, by = "fun")
```

```{r ebirdst-hidden, echo = FALSE}
read_csv("output/ebirdst_py-vs-r.csv")
```
