---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  error = FALSE
)
```

# gdal-summarize

The goal of `gdal-summarize.py` is to summarize raster data across layers. There are two common use cases for this tool. The first is calculating a cell-wise summary across the bands of a raster file (e.g. a GeoTIFF). For example, given a multi-band input GeoTIFF file `input.tif`, to calculate the cell-wise sum of the first three bands use:

```{bash, eval=FALSE}
gdal-summarize.py input.tif --bands 1 2 3 --outfile output.tif
```

Alternatively, the compute the cell-wise sum across multiple GeoTIFF files (`input1.tif`, `input2.tif`, and `input3.tif`) use:

```{bash, eval=FALSE}
gdal-summarize.py input1.tif input2.tif input3.tif --outfile output.tif
```

If these input files have multiple bands, the default behavior is to summarize the across the **first** band of each file; however, the `--bands` argument can override this behavior:

```{bash, eval=FALSE}
# summarize across the second band of each file
gdal-summarize.py input1.tif input2.tif input3.tif --bands 2 --outfile output.tif
# summarize across band 1 of input1.tif and band 2 of input2.tif
gdal-summarize.py input1.tif input2.tif --bands 1 2 --outfile output.tif
```


For usage examples of `gdal-summarize.py` and a comparison with R, look at [this document](https://github.com/mstrimas/gdal-summarize/blob/master/raster-summarize.md) on the GitHub repo.  The code for the excellent [`gdal_calc.py`](https://gdal.org/programs/gdal_calc.html) served as the starting point for `gdal-summarize.py`.

## Summary Functions

The default behavior is to perform a cell-wise sum; however, other summary functions are available via the `--function` argument:

- `sum`: cell-wise sum across layers.
- `mean`: cell-wise mean across layers.
- `count`: count the number layers with non-negative value for each cell.
- `richness`: count the number of layers with positive values for each cell.

## Example dataset

This repository contains two example datasets for testing `gdal-summarize.py`. The **small example**, consists of two GeoTIFF files (`data/small1.tif` and `data/small2.tif`) each of which has two bands with dimensions 3x3. The files are filled with count data (i.e. non-negative integers) and contain both zeros and missing values. These files are meant to be small enough that the correct results can be manually calculated to help understand how the functions are working. The data in the first file looks like this:

```{r small, echo=FALSE}
r_small <- raster::stack("data/small1.tif")
rasterVis::levelplot(r_small, 
                     col.regions = viridis::viridis,
                     xlab = NULL, ylab = NULL, 
                     scales = list(drasw = FALSE),
                     names.attr = c("Band 1", "Band 2"))
```

The **large example**, consists of two GeoTIFF files (`data/large1.tif` and `data/large2.tif`) each of which has 9 bands with dimensions 250x250. The files are filled with simulated species occupancy (i.e. real numbers between 0 and 1) and contain both zeros and missing values. These files were generated using the `simulate_species()` function from the [`prioritizr` R package](https://prioritizr.net/). The data in the first file looks like this:

```{r large, echo=FALSE}
r <- raster::stack("data/large1.tif")
rasterVis::levelplot(r, 
                     col.regions = viridis::viridis,
                     xlab = NULL, ylab = NULL, 
                     scales = list(draw = FALSE),
                     names.attr = paste("Band", seq.int(raster::nlayers(r))))
```

## Dependencies

`gdal-summarize.py` uses the GDAL Python 3 bindings and therefore requires [GDAL](https://gdal.org/) be installed on your system. On Mac OS, this can be accomplished using [homebrew](https://brew.sh/) with:

```{bash, eval=FALSE}
brew install gdal
```

## Getting Help

Use `gdal-summarize.py -h` to get help and usage guidelines for this tool:

```{bash, eval=FALSE}
usage: gdal-summarize.py [-h] --outfile OUTFILE [--bands BANDS [BANDS ...]]
                         [--function {mean,sum,count,richness}]
                         [--block_size BLOCK_SIZE BLOCK_SIZE] [--nrows NROWS]
                         [--quiet] [--overwrite]
                         files [files ...]

Summarize a set of rasters layers.

positional arguments:
  files                 input raster file(s)

optional arguments:
  -h, --help            show this help message and exit
  --outfile OUTFILE, -o OUTFILE
                        output raster
  --bands BANDS [BANDS ...], -b BANDS [BANDS ...]
                        bands to summarize. single file: bands in this file to
                        summarize (default all bands); multiple files: bands
                        in corresponding files to summarize (default = 1)
  --function {mean,sum,count,richness}, -f {mean,sum,count,richness}
                        summarization function (default = 'sum')
  --block_size BLOCK_SIZE BLOCK_SIZE, -s BLOCK_SIZE BLOCK_SIZE
                        x and y dimensions of blocks to process (default based
                        on input)
  --nrows NROWS, -n NROWS
                        number of rows to process in a single block
                        (block_size ignored if provided)
  --quiet, -q           supress messages
  --overwrite, -w       overwrite existing file
```
